The Evolution of Phenotypically Invariant Genetic Networks

### 2018-01-20 kalman_walk_paper.tex
minor edits
some rewrite of refs

%\plr{Additional ideas to consider adding:}
%\begin{itemize}
%    \item speciation literature
    %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
    %\item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
    %\item hybrid vigor (need to do calculation)
 %   \item discuss linearity, linearization, and canalization in introduction
    %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
 %   \item ``On the origin of species not by means of natural selection''
%\end{itemize}

%\plri{Note: in the \LaTeX source I'm putting in semantic linebreaks, so it's easy to edit and move around phrases and ideas.}

%\plri{Need to come up with a consistent term for ``the $A_{ij}$''s. -- ``regulatory coefficients''? ``genotype''?}

@@

It is not a new observation that there is often more than one way to do the same thing, or that speciation can be the result of neutral and quasi-neutral processes \citep{wright1932roles, nei1983models, barton1989divergence, barton2001role, gavrilets2004fitness}.
%Incompatibilities between distinct genotypes encoding identical phenotypes can
%even persist stably within a species \citep{phillips1996maintenance}.
For instance, \citet{tulchinsky2014hybrid}
%(and similarly \citet{khatri2015simple})
looked for hybrid incompatibility
following regulatory sequence drift,
and \citet{porter2002speciation} studied speciation in populations expressing a simple regulatory cascade.
Both found rapid speciation following directional selection,
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.
%REFERENCE: Barton's paper (PETER). \citep{barton2010natural}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}

@@

\paragraph{Haldane's rule}
This model naturally predicts
Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex}.
For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
A male whose pair of haplotypes is 
$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
if dosage compensation upregulates heterogametes by a factor of 2 relative to homogametes,
% as in \emph{Drosophila}
while a female homozygous for the haplotype
$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$, has phenotype determined by
$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
An $F_1$ male offspring of these two will have its phenotype determined by
$\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
If both genes resided on the autosomes this system would only be possible in an $F_2$ cross.
More generally, 
if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids. 
In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, the observation that $F_2$ hybrids will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
%and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
%\plr{what are they?}
%\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$. In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.

@@

Consistent with empirical observation of hybrid breakdown \citep[e.g.][]{plotner2017chlorosis}, 
we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
% since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
%\plr{refer to Turelli here}
% In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
Another natural consequence of the model is Haldane's rule,
that if only one $F_1$ hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
This occurs because if the genes underlying a regulatory network are distributed among both autosomes and the sex chromosome, 
then heterogametic $F_1$s show variation similar to that seen in $F_2$ hybrids.

@@

\paragraph{On the origin of species not by means of natural selection?}
% this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.
% other mechanisms also probly important
% but this one may be strong since happens in all networks at once
Although, as classically formulated, 
the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary},
and plausible mechanisms for the origin of Dobzhansky--Muller incompatibilities by neutral genetic drift have been proposed \citep{lynch2000origin} 
and under stabilizing selection \citep{fierst},
previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
(Also note that the mechanism underlying hybrid inviability here 
resembles the Dobzhansky-Muller model,
but is more similar to the ``pathway model'' of \citet{lindtke2015genetic}.)
Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearance of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
This, in light of the few known incompatibilities, has lead some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence} 
or a consequence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual}.
In contrast, the model we develop here suggests that even under strictly neutral conditions, 
system drift can lead to speciation, at a rate fast enough 
to play a substantial role in species formation across the tree of life. 
Our results show that hybrids, under neutral processes, break down as a function of genetic distance,
and so may, in part, explain broad patterns
such as the relationship between molecular divergence and genetic isolation 
% across diverse taxa and ecologies 
seen by \citet{roux2016shedding}, 
and the clocklike speciation patterns observed by \citet{hedges2015tree}.
% (as diversification is not observed to be accelerated following mass extinctions).

@@

\section*{Acknowledgements}
We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren, and Hossein Asgharian for valuable discussion. We would also like to thank Nick Barton, Sarah Signor and Todd Parsonsfor helpful comments on the manuscript.
%
%\plr{Josh: please fix capitalization in the references.  To do this, anything that needs to be capitalized gets put in curly braces, which tells bibtex to render the capitalization 'as-is'.}

>>>>

@@

% wright1932roles mentions possibility of equivalent combinations
% wright1934evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
% barton1989divergence not available but likely following on from barton1986maintenance analyzing multiple optima in a similar model
% nei1983models simulates from simple stepwise mut model on one or two loci (without interactions) where incompat betw diff alleles increases with number of steps betw them
% sved1981twosex mathematically analyzes and (in another paper) simulates speciation in a model of male + female quantitative traits indep'ly evolving with attraction between matching traits
% barton2010natural has a good historical review of natural selection and speciation; then analyzes a two-niche levene model
% gavrilets2004fitness talks a lot about quasineutral paths in holey fitness landscapes
% phillips1996maintenance shows that subpop structure can stably maintain equivalent but incompat genotypes
% tulchinsky2014hybrid simulates a single promoter region with a bioenergentic model with directional and stabilizing selection, found incompat in both cases
% porter2002speciation simulated speciation under a simple model of a regulatory cascade
%%% Fisher's geom model: (incompat due to segregation variance)
% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model 
% fraisse2016genetics looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations
% chevin2010niche looks at incompats after adaptation in Fisher's geom model and how it depends on dimension
% simon2017coadapted follows up on the Fisher's geom model predictions, also comparing to some real data

It is not a new observation that there is often more than one way to do the same thing, 
or that speciation can be the result of (nearly) neutral processes.
This has been analyzed for traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1934evolution,barton1986maintenance,barton2001role}
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
\citet{phillips1996maintenance} showed that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes.
Recent work has shown that Fisher's geometric model of stabilizing selection
predicts many general empirical patterns \citep{fraisse2016genetics,simon2017coadapted}.
We use similar tools in the second half of this paper,
differing in that we have a submanifold of optima rather than a single point.
Using simulations, \citet{tulchinsky2014hybrid}
%(and similarly \citet{khatri2015simple})
looked for hybrid incompatibility
following regulatory sequence drift,
and \citet{porter2002speciation} studied speciation in populations expressing a simple regulatory cascade.
Both found rapid speciation following directional selection,
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}

@@

\paragraph{Haldane's rule}
This model naturally predicts
Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex}.
For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
A male whose pair of haplotypes is 
$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
if dosage compensation upregulates heterogametes by a factor of 2 relative to homogametes,
% as in \emph{Drosophila}
while a female homozygous for the haplotype
$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$, has phenotype determined by
$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
An $F_1$ male offspring of these two will have its phenotype determined by
$\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
If both genes resided on the autosomes this system would only be possible in an $F_2$ cross.
More generally, 
if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids. 
In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, the observation that $F_2$ hybrids will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
\plr{Explain that the effects on fitness are dominant.}
%and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
%\plr{what are they?}
%\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$. In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.

@@

Consistent with empirical observation of hybrid breakdown \citep[e.g.][]{plotner2017chlorosis}, 
we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
% since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
%\plr{refer to Turelli here}
% In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
Another natural consequence of the model is Haldane's rule,
that if only one $F_1$ hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
This occurs because if the genes underlying a regulatory network are distributed among both autosomes and the sex chromosome, 
then heterogametic $F_1$s show variation similar to that seen in $F_2$ hybrids.
\plr{cite fisher's model stuff}

@@

\paragraph{On the origin of species not by means of natural selection?}
% this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.
% other mechanisms also probably important
% but this one may be strong since happens in all networks at once
Although, as classically formulated, 
the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary},
and plausible mechanisms for the origin of Dobzhansky--Muller incompatibilities by neutral genetic drift have been proposed \citep{lynch2000origin} 
\plr{what}
and under stabilizing selection \citep{fierst},
previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
(Also note that the mechanism underlying hybrid inviability here 
resembles the Dobzhansky-Muller model,
but is more similar to the ``pathway model'' of \citet{lindtke2015genetic}.)
Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearance of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
This, in light of the few known incompatibilities, has lead some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence} 
or a consequence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual}.
In contrast, the model we develop here suggests that even under strictly neutral conditions, 
system drift can lead to speciation, at a rate fast enough 
to play a substantial role in species formation across the tree of life. 
Our results show that hybrids, under neutral processes, break down as a function of genetic distance,
and so may, in part, explain broad patterns
such as the relationship between molecular divergence and genetic isolation 
% across diverse taxa and ecologies 
seen by \citet{roux2016shedding}, 
and the clocklike speciation patterns observed by \citet{hedges2015tree}.
% (as diversification is not observed to be accelerated following mass extinctions).

@@

\section*{Acknowledgements}
We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren, and Hossein Asgharian for valuable discussion. We would also like to thank Nick Barton, Sarah Signor and Todd Parsons for helpful comments on the manuscript.
%
%\plr{Josh: please fix capitalization in the references.  To do this, anything that needs to be capitalized gets put in curly braces, which tells bibtex to render the capitalization 'as-is'.}
